Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems

### 2023-07-10 rmd/manuscript.Rmd
edit: proof correction
edit/correct in response to ecosphere's editorial office

Thus, our definition accounts for the influence of omnivory. As a result, the FCL in a single patch can change due to the P's degree of omnivory and/or species extinction. We defined the global FCL $\bar{\tau}$ across patches $N_p$ and time $T$ as $\bar{\tau} = \frac{\sum_x \sum_t \tau_{x,t}}{N_p T}$ (i.e., average over space and time).

We also calculated the proportion of the patch state ("No species", "B", "B + C", "B + P", and "B + C + P") at each time step to infer mechanisms that drive changes in FCL. The time-averaged proportion of patch state $\bar{\psi}$ was obtained as $\bar{\psi} = \frac{q_t}{N_p T}$, where $q_t$ is the number of patches in a given patch state at time step $t$.

@@

Prior to the main simulation, we performed one-patch simulations to explore the predation parameter space ($a_{ij}$ and $h_{ij}$) that allows stable coexistence of the three species without disturbance and dispersal (see **Supporting Information**). In this simulation, we assumed $h_{ij} = h$ to reduce the dimension of parameter space. This analysis revealed that the stable coexistence of the three species is possible with certain combinations of predation parameters. Within this parameter space, we used three parameter combinations to represent distinctive food webs with varied levels of omnivory: "chain (no omnivory)" $(a_{BC}, a_{BP}, a_{CP}, h) = (0.5, 0, 0.0025, 0.5)$, "weak omnivory" ($0.5, 0.02, 0.025, 0.5$), and "strong omnivory" ($0.5, 0.04, 0.0025, 0.75$). We used $s = 1$ except for the chain scenario with $s = 0$. Without the loss of generality, we assumed $e_{ij} = e = 1$.

@@

Ecosystem geometry interacted with disturbance to dictate FCL in branching ecosystems (**Figures 3 and 4**). In the presence of disturbance, both geometric features (size and complexity) had noticeable influences on FCL. As disturbance frequency increased, the relationships between FCL and ecosystem geometry became clearer and positive. This interactive effect emerged because the sensitivity of FCL to disturbance depended on ecosystem geometry. Although increased disturbance frequency reduced overall FCL, smaller and/or simpler ecosystems exhibited greater vulnerability (**Figures 3 and 4**). Therefore, disturbance interacted with ecosystem geometry to produce the interdependence of these factors in determining FCL. Yet, the interactive effects were not observed with low disturbance intensity (**Figures S3 and S4**); this suggests that disturbance intensity must be strong enough to be influential.

With no disturbance, the effects of ecosystem geometry were highly context dependent (**Figure 2**). In the chain scenario (no omnivory) with low productivity, ecosystem size had a positive effect on FCL (**Figure 3**), while ecosystem complexity had a negative influence (**Figure 4**). These opposing influences of ecosystem size and complexity were associated with their contrasting relationships with the total carrying capacity (i.e., carrying capacity summed across all patches): ecosystem size increased the total carrying capacity, whereas ecosystem complexity decreased it (**Figure S8**). However, when considering some levels of omnivory, these patterns characteristic of the chain scenario diminished. With weak omnivory (the second rows of **Figure 3 and 4**), ecosystem complexity had consistent positive effects on FCL, except the scenario with low productivity and short dispersal (**Figure 4**). In contrast, the effect of ecosystem size was more productivity dependent (**Figure 4**). Neither of ecosystem size and complexity influenced FCL with strong omnivory.

State transitions largely depended on motif structure and shed light on how FCL changes occurred. In the chain motif, an additive mechanism (i.e., the addition of higher trophic species) was always responsible for the changes in FCL. With disturbance, state transition occurs from "B" to "B + C" (low productivity) or "B + C" to "B + C + P" (high productivity) as ecosystem size or complexity increases (**Figures 5 and 6**). Exceptionally, the state transition of "B + C + P" $\rightarrow$ "B + C" occurred as ecosystem complexity increased with no disturbance and low productivity, a scenario in which ecosystem complexity decreased FCL.

@@

In the chain motif, productivity greatly increased the overall FCL (**Figure S5**), thereby shifting the upper bound of FCL (**Figures 3 and 4**). In contrast, increased productivity decreased FCL with weak and strong omnivory (**Figures 3, 4, and S5**). This negative effect of increased productivity was apparent in smaller and/or simpler ecosystems, steepening the slopes between FCL and ecosystem geometry (**Figures 3 and 4**). State transitions associated with increased productivity depended on the motif structure. In the chain motif, the transition from "B + C" to "B + C + P" was dominant regardless of disturbance frequency (**Figure S6**). The introduction of omnivory allowed the transition between "B + P" and "B + C + P". With weak omnivory, the transition of "B + C + P" $\rightarrow$ "B + P" was prevalent and shortened FCL across disturbance regimes (**Figure S6**). However, this transition was marginal in the strong omnivory motif. Instead, productivity significantly increased the proportion of the "No species" state (**Figure S6**) in disturbed scenarios, especially in small and/or simple ecosystems (**Figures 5 and 6**).

Dispersal distance had a positive influence on FCL regardless of simulation scenarios (**Figure S5**). State transitions occurred through an additive (chain) or insertion mechanism (omnivory). In the chain motif, the transitions were either "B" $\rightarrow$ "B + C" or "B + C" $\rightarrow$ "B + C+ P", depending on disturbance frequency (**Figure S7**). In the omnivory motifs, the predominant transition was from "B + P" to "B + C + P" (**Figure S7**). However, this direct effect was weak in comparison to other factors with no qualitative alteration to the relationships between FCL and ecosystem geometry (**Figures 3 and 4**).

@@

The most significant finding in this study is the synergistic effects between ecosystem geometry and disturbance. Our no-disturbance scenario can be viewed as habitat networks having nearly homogeneous environments, in which long food chains were generally supported independently of ecosystem size and complexity (except the chain motif with low productivity). However, as the frequency of intense disturbance increases, small and/or simple networks become unable to maintain long food chains. Here, increased ecosystem size and complexity may give greater resistance to disturbance through different mechanisms. Ecosystem size increases the "total amount" of disturbance refugia, from which dispersing propagules recolonize impacted habitat patches. In the meantime, ecosystem complexity may increase the connectivity to disturbance refugia. Branching shortens the mean distance between patches and, therefore, may enhance immigration from less disturbed patches (**Figure S10**). The proportion of disturbance refugia may also have increased in more branched networks. However, we did not observe a clear proportional increase in disturbance refugia with increasing branching probability (**Figure 1B**). Collectively, each of the geometric features likely supports long food chains through distinct mechanisms, which may also explain the different patterns of state transitions (additive vs. insertion mechanisms) [@postProximateStructuralMechanisms2007]. We must note that, however, the effects of ecosystem size and complexity were limited to small to intermediate levels. In large or complex ecosystems, dispersal propagules may not be a limiting factor of FCL as the supply is plenty; therefore, productivity or omnivory more strongly regulated FCL (**Figures 2 and 3**) as predicted by previous theory [@takimotoEffectsProductivityDisturbance2012]

@@

Another interesting result of our simulation was the interactive influence of omnivory. In particular, with no disturbance and low productivity, the FCL of the chain motif had contrasting relationships with ecosystem size (positive) and complexity (negative). It is evident from the one-patch simulation that the chain motif is highly sensitive to local carrying capacity (**Figures S1 and S2**); therefore, FCL followed the changes in the total carrying capacity along the gradients of ecosystem size and complexity. These relationships diminished in omnivory motifs as the dependence of FCL on local carrying capacity weakened or reversed (**Figures S1 and S2**). Importantly, however, disturbance overwhelmed the role of omnivory in modulating the FCL-geometry relationships: as the frequency of intense disturbance increases, the positive relationships between FCL and ecosystem geometry emerged regardless of omnivory levels. This result could explain the consistent support for the ecosystem size hypothesis in various ecosystems [@takimotoEnvironmentalDeterminantsFoodchain2013] and raises the possibility of ecosystem complexity as an important control of FCL in branching networks. Our results also imply that if disturbance is strong and frequent, the effects of ecosystem geometry are robust to food web reorganization, which is often associated with species extirpations or introductions. The strength of omnivory in these novel communities can be increased or decreased, depending on the introduced species and the extant species pool or community. For example, non-indigenous fishes in the Eastern Mediterranean Sea outcompete native species for high-quality prey items, causing native species to increase their consumption of non-preferred prey items [@fanelliDepictingNovelEastern2015]. Yet, such substantial food web reorganization may not alter qualitative relationships between FCL and ecosystem geometry.

The effect of productivity was motif-dependent, supporting previous theoretical predictions [@postProximateStructuralMechanisms2007; @kondohFoodchainLengthAdaptive2009; @wardMechanisticTheoryAquatic2017]. While productivity clearly increased FCL in the chain motif, the productivity effect was negative in omnivory motifs. Intraguild predation plays an important role in explaining this context dependency. In the high productivity scenario with omnivory, the increased abundance of the basal species enhances the intraguild predator. This may subsequently facilitate the local exclusion of the intraguild prey through increased predation pressure if the attack rate is sufficiently high [@postProximateStructuralMechanisms2007]. Indeed, in the weak omnivory motif (the high value of the attack rate of P on C), we observed the state transition from "B + C + P" to "B + P" with shortened FCL regardless of disturbance frequency. However, this was not the case in the strong omnivory motif; the increased proportion of the "No species" state strongly impacted the minimum FCL, and this transition was mediated by disturbance. This result suggests that, although the strong omnivory motif is innately stable with high productivity (see one-patch simulations; **Figures S1 and S2**), external disturbance undermines the sensitive balance of predator-prey interactions with the abrupt increase of "No species" state, which must be countered by immigration from less disturbed habitats to maintain food chains.

@@

**Figure 2** Summary of our simulation results. Black text in each node indicates the potential factors that may influence the association between food chain length and ecosystem geometry, and black text on arrows indicates simulated scenarios for a given factor. Blue text denotes the resultant association between food chain length and ecosystem geometry. Panel (A) indicates the effect of ecosystem size. Panel (B) indicates the effect of ecosystem complexity.

@@

**Figure 5** Relationships between the proportion of each patch state $\bar{\psi}$ (colors indicate different patch states) and ecosystem size with high disturbance intensity $\mu_m = 0.8$. The proportion of each patch state $\bar{\psi}$ is averaged over time for each simulation replicate. Lines indicate loess curves fitted to $500$ simulation replicates, and shades are associated 95% confidence intervals. Panels indicate different food web motifs. In each panel, rows distinguish disturbance probability $p_m$, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities.

**Figure 6** Relationships between the proportion of each patch state $\bar{\psi}$ (colors indicate different patch states) and ecosystem complexity with high disturbance intensity $\mu_m = 0.8$. The proportion of each patch state $\bar{\psi}$ is averaged over time for each simulation replicate. Lines indicate loess curves fitted to 500 simulation replicates, and shades are associated with 95% confidence intervals. Panels indicate different food web motifs. In each panel, rows distinguish disturbance probability $p_m$, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities.

>>>>

Thus, our definition accounts for the influence of omnivory. As a result, the FCL in a single patch can change due to the P's degree of omnivory and/or species extinction. We defined the global FCL $\bar{\tau}$ across patches $N_p$ and time $T$ as $\bar{\tau} = \sum_x \sum_t \tau_{x,t}/(N_p T)$ (i.e., average over space and time).

We also calculated the proportion of the patch state ("No species", "B", "B + C", "B + P", and "B + C + P") at each time step to infer mechanisms that drive changes in FCL. The time-averaged proportion of patch state $\bar{\psi}$ was obtained as $\bar{\psi} = q_t/(N_p T)$, where $q_t$ is the number of patches in a given patch state at time step $t$.

@@

Prior to the main simulation, we performed one-patch simulations to explore the predation parameter space ($a_{ij}$ and $h_{ij}$) that allows stable coexistence of the three species without disturbance and dispersal (see **Appendix S1: Supplementary Text**). In this simulation, we assumed $h_{ij} = h$ to reduce the dimension of parameter space. This analysis revealed that the stable coexistence of the three species is possible with certain combinations of predation parameters. Within this parameter space, we used three parameter combinations to represent distinctive food webs with varied levels of omnivory: "chain (no omnivory)" $(a_{BC}, a_{BP}, a_{CP}, h) = (0.5, 0, 0.0025, 0.5)$, "weak omnivory" ($0.5, 0.02, 0.025, 0.5$), and "strong omnivory" ($0.5, 0.04, 0.0025, 0.75$). We used $s = 1$ except for the chain scenario with $s = 0$. Without the loss of generality, we assumed $e_{ij} = e = 1$.

@@

Ecosystem geometry interacted with disturbance to dictate FCL in branching ecosystems (**Figures 3 and 4**). In the presence of disturbance, both geometric features (size and complexity) had noticeable influences on FCL. As disturbance frequency increased, the relationships between FCL and ecosystem geometry became clearer and positive. This interactive effect emerged because the sensitivity of FCL to disturbance depended on ecosystem geometry. Although increased disturbance frequency reduced overall FCL, smaller and/or simpler ecosystems exhibited greater vulnerability (**Figures 3 and 4**). Therefore, disturbance interacted with ecosystem geometry to produce the interdependence of these factors in determining FCL. Yet, the interactive effects were not observed with low disturbance intensity (**Appendix S1: Figures S3 and S4**); this suggests that disturbance intensity must be strong enough to be influential.

With no disturbance, the effects of ecosystem geometry were highly context dependent (**Figure 2**). In the chain scenario (no omnivory) with low productivity, ecosystem size had a positive effect on FCL (**Figure 3**), while ecosystem complexity had a negative influence (**Figure 4**). These opposing influences of ecosystem size and complexity were associated with their contrasting relationships with the total carrying capacity (i.e., carrying capacity summed across all patches): ecosystem size increased the total carrying capacity, whereas ecosystem complexity decreased it (**Appendix S1: Supplementary Text8**). However, when considering some levels of omnivory, these patterns characteristic of the chain scenario diminished. With weak omnivory (the second rows of **Figure 3 and 4**), ecosystem complexity had consistent positive effects on FCL, except the scenario with low productivity and short dispersal (**Figure 4**). In contrast, the effect of ecosystem size was more productivity dependent (**Figure 4**). Neither of ecosystem size and complexity influenced FCL with strong omnivory.

State transitions largely depended on motif structure and shed light on how FCL changes occurred. In the chain motif, an additive mechanism (i.e., the addition of higher trophic species) was always responsible for the changes in FCL. With disturbance, state transition occurs from "B" to "B + C" (low productivity) or "B + C" to "B + C + P" (high productivity) as ecosystem size or complexity increases (**Figures 5 and 8**). Exceptionally, the state transition of "B + C + P" $\rightarrow$ "B + C" occurred as ecosystem complexity increased with no disturbance and low productivity, a scenario in which ecosystem complexity decreased FCL.

@@

In the chain motif, productivity greatly increased the overall FCL (**Appendix S1: Figure S5**), thereby shifting the upper bound of FCL (**Figures 3 and 4**). In contrast, increased productivity decreased FCL with weak and strong omnivory (**Figures 3, 4, and Appendix S1: Figure S5**). This negative effect of increased productivity was apparent in smaller and/or simpler ecosystems, steepening the slopes between FCL and ecosystem geometry (**Figures 3 and 4**). State transitions associated with increased productivity depended on the motif structure. In the chain motif, the transition from "B + C" to "B + C + P" was dominant regardless of disturbance frequency (**Appendix S1: Figure S6**). The introduction of omnivory allowed the transition between "B + P" and "B + C + P". With weak omnivory, the transition of "B + C + P" $\rightarrow$ "B + P" was prevalent and shortened FCL across disturbance regimes (**Appendix S1: Figure S6**). However, this transition was marginal in the strong omnivory motif. Instead, productivity significantly increased the proportion of the "No species" state (**Appendix S1: Figure S6**) in disturbed scenarios, especially in small and/or simple ecosystems (**Figures 7 and 10**).

Dispersal distance had a positive influence on FCL regardless of simulation scenarios (**Appendix S1: Figure S5**). State transitions occurred through an additive (chain) or insertion mechanism (omnivory). In the chain motif, the transitions were either "B" $\rightarrow$ "B + C" or "B + C" $\rightarrow$ "B + C+ P", depending on disturbance frequency (**Appendix S1: Figure S7**). In the omnivory motifs, the predominant transition was from "B + P" to "B + C + P" (**Appendix S1: Figure S7**). However, this direct effect was weak in comparison to other factors with no qualitative alteration to the relationships between FCL and ecosystem geometry (**Figures 3 and 4**).

@@

The most significant finding in this study is the synergistic effects between ecosystem geometry and disturbance. Our no-disturbance scenario can be viewed as habitat networks having nearly homogeneous environments, in which long food chains were generally supported independently of ecosystem size and complexity (except the chain motif with low productivity). However, as the frequency of intense disturbance increases, small and/or simple networks become unable to maintain long food chains. Here, increased ecosystem size and complexity may give greater resistance to disturbance through different mechanisms. Ecosystem size increases the "total amount" of disturbance refugia, from which dispersing propagules recolonize impacted habitat patches. In the meantime, ecosystem complexity may increase the connectivity to disturbance refugia. Branching shortens the mean distance between patches and, therefore, may enhance immigration from less disturbed patches (**Appendix S1: Figure S8**). The proportion of disturbance refugia may also have increased in more branched networks. However, we did not observe a clear proportional increase in disturbance refugia with increasing branching probability (**Figure 1B**). Collectively, each of the geometric features likely supports long food chains through distinct mechanisms, which may also explain the different patterns of state transitions (additive vs. insertion mechanisms) [@postProximateStructuralMechanisms2007]. We must note that, however, the effects of ecosystem size and complexity were limited to small to intermediate levels. In large or complex ecosystems, dispersal propagules may not be a limiting factor of FCL as the supply is plenty; therefore, productivity or omnivory more strongly regulated FCL (**Figures 2 and 3**) as predicted by previous theory [@takimotoEffectsProductivityDisturbance2012]

@@

Another interesting result of our simulation was the interactive influence of omnivory. In particular, with no disturbance and low productivity, the FCL of the chain motif had contrasting relationships with ecosystem size (positive) and complexity (negative). It is evident from the one-patch simulation that the chain motif is highly sensitive to local carrying capacity (**Appendix S1: Figures S1 and S2**); therefore, FCL followed the changes in the total carrying capacity along the gradients of ecosystem size and complexity. These relationships diminished in omnivory motifs as the dependence of FCL on local carrying capacity weakened or reversed (**Appendix S1: Figures S1 and S2**). Importantly, however, disturbance overwhelmed the role of omnivory in modulating the FCL-geometry relationships: as the frequency of intense disturbance increases, the positive relationships between FCL and ecosystem geometry emerged regardless of omnivory levels. This result could explain the consistent support for the ecosystem size hypothesis in various ecosystems [@takimotoEnvironmentalDeterminantsFoodchain2013] and raises the possibility of ecosystem complexity as an important control of FCL in branching networks. Our results also imply that if disturbance is strong and frequent, the effects of ecosystem geometry are robust to food web reorganization, which is often associated with species extirpations or introductions. The strength of omnivory in these novel communities can be increased or decreased, depending on the introduced species and the extant species pool or community. For example, non-indigenous fishes in the Eastern Mediterranean Sea outcompete native species for high-quality prey items, causing native species to increase their consumption of non-preferred prey items [@fanelliDepictingNovelEastern2015]. Yet, such substantial food web reorganization may not alter qualitative relationships between FCL and ecosystem geometry.

The effect of productivity was motif-dependent, supporting previous theoretical predictions [@postProximateStructuralMechanisms2007; @kondohFoodchainLengthAdaptive2009; @wardMechanisticTheoryAquatic2017]. While productivity clearly increased FCL in the chain motif, the productivity effect was negative in omnivory motifs. Intraguild predation plays an important role in explaining this context dependency. In the high productivity scenario with omnivory, the increased abundance of the basal species enhances the intraguild predator. This may subsequently facilitate the local exclusion of the intraguild prey through increased predation pressure if the attack rate is sufficiently high [@postProximateStructuralMechanisms2007]. Indeed, in the weak omnivory motif (the high value of the attack rate of P on C), we observed the state transition from "B + C + P" to "B + P" with shortened FCL regardless of disturbance frequency. However, this was not the case in the strong omnivory motif; the increased proportion of the "No species" state strongly impacted the minimum FCL, and this transition was mediated by disturbance. This result suggests that, although the strong omnivory motif is innately stable with high productivity (see one-patch simulations; **Appendix S1: Figures S1 and S2**), external disturbance undermines the sensitive balance of predator-prey interactions with the abrupt increase of "No species" state, which must be countered by immigration from less disturbed habitats to maintain food chains.

@@

**Figure 2** Summary of our simulation results. Black text in each node indicates the potential factors that may influence the association between food chain length and ecosystem geometry, and black text on arrows indicates simulated scenarios for a given factor. Blue signs at the terminal nodes denote the resultant association between food chain length and ecosystem geometry (+: positive, -: negative, 0: no effect, (0): very weak effect). Panel (A) indicates the effect of ecosystem size. Panel (B) indicates the effect of ecosystem complexity.

@@

**Figure 5** Relationships between the proportion of each patch state $\bar{\psi}$ (colors indicate different patch states) and ecosystem size under the chain motif scenario. Disturbance intensity is $\mu_m = 0.8$. The proportion of each patch state $\bar{\psi}$ is averaged over time for each simulation replicate. Lines indicate loess curves fitted to $500$ simulation replicates, and shades are associated 95% confidence intervals.Rows distinguish disturbance probability $p_m$, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities.

**Figure 6** Relationships between the proportion of each patch state $\bar{\psi}$ (colors indicate different patch states) and ecosystem size under the weak omnivory scenario. Additional specifics can be found in Figure 5.

**Figure 7** Relationships between the proportion of each patch state $\bar{\psi}$ (colors indicate different patch states) and ecosystem size under the strong omnivory scenario. Additional specifics can be found in Figure 5.

**Figure 8** Relationships between the proportion of each patch state $\bar{\psi}$ (colors indicate different patch states) and ecosystem complexity under the chain motif scenario. Disturbance intensity is $\mu_m = 0.8$. The proportion of each patch state $\bar{\psi}$ is averaged over time for each simulation replicate. Lines indicate loess curves fitted to 500 simulation replicates, and shades are associated with 95% confidence intervals. Rows distinguish disturbance probability $p_m$, and columns indicate different dispersal and productivity scenarios. Different line colors indicate different disturbance probabilities.

**Figure 9** Relationships between the proportion of each patch state $\bar{\psi}$ (colors indicate different patch states) and ecosystem complexity under the weak omnivory scenario. Disturbance intensity is $\mu_m = 0.8$. Additional specifics can be found in Figure 8.

**Figure 10** Relationships between the proportion of each patch state $\bar{\psi}$ (colors indicate different patch states) and ecosystem complexity under the strong omnivory scenario. Disturbance intensity is $\mu_m = 0.8$. Additional specifics can be found in Figure 8.
